Towards Automated Refactoring of Code Clones in Object-Oriented Programming Languages\\- Work in Progress -

### 2019-07-01 sattose/paper.tex
Spelling mainly

\begin{abstract}
Duplication in source code can have a major negative impact on the maintainability of source code, as it creates implicit dependencies between fragments of code. Such implicit dependencies often cause bugs and increases maintenance efforts. In this study, we look into the opportunities to automatically refactor these duplication problems for object-oriented programming languages. We propose a method to detect clones that are suitable for refactoring. This method focuses on the context and scope of clones, ensuring our refactoring improves the design and does not create side effects.

@@

A tool named Aries~\cite{higo2004aries, higo2008metric} focuses on the detection of refactorable clones. They do this based on the relation between clone instances through inheritance, similar to Fontana et al.~\cite{fontana2012duplicated}. This tool only proposes a refactoring opportunity, and does not provide help in the process of applying the refactoring.

@@

In this section we discuss the shortcomings of the different clone type definitions. Because of these shortcomings, clones found by these definitions are often found to require additional judgement whether they should and can be refactored.

@@

Type 1 clones are by most tools implemented as textual equality between code fragments (except for whitespace and comments). Although textually equal, method calls can still refer to different methods, type declarations can still refer to different types and variables can be of a different type. In such cases refactoring opportunities could be invalidated. This can make type 1 clones less suitable for refactoring purposes, as they require additional judgment regarding the refactorability of such a clone. When aiming to automatically refactor clones, applying refactorings to clones which might refer to different methods, types and variables is bound to be error prone and result in a uncompilable project or a difference in functionality.

Because of this, not all type 1 clones may be subject to refactoring. In section \label{chap:type1rclones} we describe an alternate approach towards detecting type 1 clones, which results in only clones that can be refactored.

@@

When looking at refactoring, type 2 clones can be difficult to refactor. For instance if we have variability in types, the code can describe operations on two completely dissimilar types. Type 2 clones do not differentiate between primitives and referenced types, which further undermines the usefulness of clones detected by this definition.

\subsection{Type 3 clones}\label{sec:type3}
Type 3 clones are \textit{copied fragments with further modification (having added, removed or changed statements)}~\cite{roy2007survey}. Detection of clones by this definition can be hard, as it may be hard to detect whether a fragment was copied in the first place if it was severely changed. Because of this, most clone detection implementations of type 3 clones work on basis of a similarity threshold~\cite{roy2008nicad,ragkhitwetsagul2019siamese,jiang2007deckard,semura2017ccfindersw}. This similarity threshold has been implemented in different ways: textual similarity (for instance using levenshtein distance)~\cite{lavoie2011automated}, token-level similarity~\cite{sajnani2016sourcerercc} or statement-level similarity~\cite{kamalpriya2017enhancing}.

Having a definition that allows for any change in code poses serious challenges on refactoring. A levenshtein distance of one can already change the meaning of a code fragment significantly, for instance if the name of a type differs by a character (and thus referring to different types).

\subsection{Refactoring-oriented clone types}\label{sec:rtypes}
To resolve the shortcomings of clone types as outlined in the previous section, we propose alternative definitions for clone types to be directed at detecting clones that can and should be refactored. We have named these clones T1R (type 1R), T2R and T3R clones. These definitions share similarities with the literature definitions, the number of each type corresponds with the clone type it is modeled after. The ``R'' stands for refactoring-oriented.

@@

To limit the negative impact of this tradeoff on the design of the system, we formalized as threshold for the variability between fragments. The formula by which this threshold is calculated is displayed in equation \ref{eq:type2r}. In this formula, \textit{different expressions} refers to the number of literals, variables and method calls that differ from other clone instances in a clone class. We divide this by the total number of tokens in the clone instance. Based on this threshold, we decide whether a clone should be considered for refactoring.

@@

\begin{itemize}
  \item \textbf{The difference in statements must bridge a gap between two clones that were valid by the original thresholds}. This entails that, different from type 3 clones, the difference in statements cannot be at the beginning or the end of a cloned block. It is rather somewhere within, as it must bridge two existent clones.
  \item \textbf{The size of the gap between two clones is limited by a threshold.} This threshold is calculated by taking the percentage of the number of statements in the gap over the number of statements that both clones that are being bridged span. This is displayed in equation \ref{eq:type3r}.
  \item \textbf{The gap may not span a partial block.} To make sure that the T3R clone can be refactored, we do not allow the gap to span a part of a block, for instance the declaration and a part of the body of a for-loop. The reason for this is that it is not possible to wrap a partially spanned block in a single conditional statement. We could however use multiple conditional blocks (one for each block spanned), but due to the detrimental effect on the design of the code (as each conditional block adds a certain complexity), we decided not to allow this for T3R clones.
\end{itemize}

@@

\subsubsection{The challenge of detecting these clones}\label{chap:challenge}
To detect each type of clone, we need to parse the FQI of all types, method calls and variables. This comes with challenges, regarding both performance and implementation. To trace the declarations of variables, methods and types, we might need to follow cross file references. The referenced types/variables/methods might even not be part of the project, but rather of an external library or the standard libraries of the programming language. All these factors need to be considered for the referenced entity to be found, on basis of which an FQI can be created.

@@

\subsection{Survey on Clone Detection Tools}
\label{ch:tool-overview}
We conducted a short survey on (recent) clone detection tools that we could use to analyze refactoring possibilities. The results of our survey are displayed in table~\ref{table:dettools}. We chose a set of tools that are open source and can analyze a popular object-oriented programming language. Next, we formulate the following four criteria by which we analyze these tools:
\begin{enumerate}
    \item \textbf{Should find clones in any context.} Some tools only find clones in specific contexts, such as only method-level clones. We want to perform an analysis on all clones in projects to get a complete overview.
\item \textbf{Finds clone classes in control projects.} We assembled a number of control projects to assess the validity of clone detection tools. On basis of this, we checked whether clone detection tools can correctly find clones in diverse contexts.
\item \textbf{Can analyse resolved symbols.} When detecting clones for refactoring purposes, it is important that clone instances can be merged. Sometimes, textual equality between code fragments does not imply that these can be merged (this is described more elaborately in Sec.~\ref{sec:type1}). Because of this, we want to use a clone detection tool that can analyze such structures.
\item \textbf{Extensive detection configuration.} We aim to exclude expressions/statements from matching (more about our rationale in section~\ref{chap:clonetypes}). To achieve this, the tool needs to be able to allow those threshold changes. This can be either through simple changes of the source code, or by using some configuration file.
\end{enumerate}

>>>>

\begin{abstract}
Duplication in source code can have a major negative impact on the maintainability of source code, as it creates implicit dependencies between fragments of code. Such implicit dependencies often cause bugs and increase maintenance efforts. In this study, we look into the opportunities to automatically refactor these duplication problems for object-oriented programming languages. We propose a method to detect clones that are suitable for refactoring. This method focuses on the context and scope of clones, ensuring our refactoring improves the design and does not create side effects.

@@

A tool named Aries~\cite{higo2004aries, higo2008metric} focuses on the detection of refactorable clones. They do this based on the relation between clone instances through inheritance, similar to Fontana et al.~\cite{fontana2012duplicated}. This tool only proposes a refactoring opportunity and does not provide help in the process of applying the refactoring.

@@

In this section, we discuss the shortcomings of the different clone type definitions. Because of these shortcomings, clones found by these definitions are often found to require additional judgment whether they should and can be refactored.

@@

Type 1 clones are by most tools implemented as textual equality between code fragments (except for whitespace and comments). Although textually equal, method calls can still refer to different methods, type declarations can still refer to different types and variables can be of a different type. In such cases, refactoring opportunities could be invalidated. This can make type 1 clones less suitable for refactoring purposes, as they require additional judgment regarding the refactorability of such a clone. When aiming to automatically refactor clones, applying refactorings to type 1 clones is bound to be error prone and can result in an uncompilable project or a difference in functionality.

Because of this, not all type 1 clones may be subject to refactoring. In section \label{chap:type1rclones} we describe an alternative approach towards detecting type 1 clones, which results in only clones that can be refactored.

@@

When looking at refactoring, type 2 clones can be difficult to refactor. For instance, if we have variability in types, the code can describe operations on two completely dissimilar types. Type 2 clones do not differentiate between primitives and reference types, which further undermines the usefulness of clones detected by this definition.

\subsection{Type 3 clones}\label{sec:type3}
Type 3 clones are \textit{copied fragments with further modification (having added, removed or changed statements)}~\cite{roy2007survey}. Detection of clones by this definition can be hard, as it may be hard to detect whether a fragment was copied in the first place if it was severely changed. Because of this, most clone detection implementations of type 3 clones work on basis of a similarity threshold~\cite{roy2008nicad,ragkhitwetsagul2019siamese,jiang2007deckard,semura2017ccfindersw}. This similarity threshold has been implemented in different ways: textual similarity (for instance using Levenshtein distance)~\cite{lavoie2011automated}, token-level similarity~\cite{sajnani2016sourcerercc} or statement-level similarity~\cite{kamalpriya2017enhancing}.

Having a definition that allows for any change in code poses serious challenges on refactoring. A Levenshtein distance of one can already change the meaning of a code fragment significantly, for instance, if the name of a type differs by a character (and thus referring to different types).

\subsection{Refactoring-oriented clone types}\label{sec:rtypes}
To resolve the shortcomings of clone types as outlined in the previous section, we propose alternative definitions for clone types directed at detecting clones that can and should be refactored. We have named these clones T1R (type 1R), T2R and T3R clones. These definitions address problems of the corresponding literature definitions. The ``R'' stands for refactoring-oriented.

@@

To limit the negative impact of this tradeoff on the design of the system, we formalized a threshold for the variability between fragments. The formula by which this threshold is calculated is displayed in equation \ref{eq:type2r}. In this formula, \textit{different expressions} refers to the number of literals, variables and method calls that differ from other clone instances in a clone class. We divide this by the total number of tokens in the clone instance. Based on this threshold, we decide whether a clone should be considered for refactoring.

@@

\begin{itemize}
  \item \textbf{The difference in statements must bridge a gap between two clones that were valid by the original thresholds}. This entails that, different from type 3 clones, the difference in statements cannot be at the beginning or the end of a cloned block. It is rather somewhere within, as it must bridge two existent clones.
  \item \textbf{The size of the gap between two clones is limited by a threshold.} This threshold is calculated by taking the percentage of the number of statements in the gap over the number of statements that both clones that are being bridged span. This is displayed in equation \ref{eq:type3r}.
  \item \textbf{The gap may not span a partial block.} To make sure that the T3R clone can be refactored, we do not allow the gap to span a part of a block, for instance, the declaration and a part of the body of a for-loop. The reason for this is that it is not possible to wrap a partially spanned block in a single conditional statement. We could, however, use multiple conditional blocks (one for each block spanned), but due to the detrimental effect on the design of the code (as each conditional block adds a certain complexity), we decided not to allow this for T3R clones.
\end{itemize}

@@

\subsubsection{The challenge of detecting these clones}\label{chap:challenge}
To detect each type of clone, we need to parse the FQI of all types, method calls, and variables. This comes with challenges, regarding both performance and implementation. To trace the declarations of variables, methods, and types, we might need to follow cross-file references. The referenced types/variables/methods might even not be part of the project, but rather of an external library or the standard libraries of the programming language. All these factors need to be considered for the referenced entity to be found, on basis of which an FQI can be created.

@@

\subsection{Survey on Clone Detection Tools}
\label{ch:tool-overview}
We conducted a short survey on (recent) clone detection tools that we could use to analyze refactoring possibilities. The results of our survey are displayed in table~\ref{table:dettools}. We chose a set of tools that are open source and can analyze a popular object-oriented programming language. Next, we formulate the following four criteria by which we analyze these tools:
\begin{enumerate}
    \item \textbf{Should find clones in any context.} Some tools only find clones in specific contexts, such as only method-level clones. We want to perform an analysis of all clones in projects to get a complete overview.
\item \textbf{Finds clone classes in control projects.} We assembled a number of control projects to assess the validity of clone detection tools. On basis of this, we checked whether clone detection tools can correctly find clones in diverse contexts.
\item \textbf{Can analyze resolved symbols.} When detecting clones for refactoring purposes, it is important that clone instances can be merged. Sometimes, textual equality between code fragments does not imply that these can be merged (this is described more elaborately in Sec.~\ref{sec:type1}). Because of this, we want to use a clone detection tool that can analyze such structures.
\item \textbf{Extensive detection configuration.} We aim to exclude expressions/statements from matching (more about our rationale in section~\ref{chap:clonetypes}). To achieve this, the tool needs to be able to allow those threshold changes. This can be either through simple changes of the source code or by using some configuration file.
\end{enumerate}
